---
title: "1 · Data preparation"
subtitle: "From a messy SPSS to a tidy dataset the rest of the site reads"
date: today
editor_options: 
  chunk_output_type: console
---

## IMPORTING SPSS HORIGIN Experiment 1B data

This is a preregistered EXP1b of HORIGIN project
Data collected at VU lab
May, 2026

```{r}
library(haven)
HO_EXP1b_raw <- read_sav("C:/Users/Zivile/Desktop/HORIGIN 1B/data/raw/HO_EXP1b_raw.sav")
View(HO_EXP1b_raw)
```

## Required packages

```{r}
install.packages("dplyr")
install.packages ("ggplot2")
install.packages("janitor")
```

## Data exclusions in line with the preregistration_test test test

We exclude participants who:
- did not finish

```{r}

library(dplyr)

HO_EXP1b_clean <- HO_EXP1b_raw|>
  filter(Finished == 1)

```

## Cleaning names

```{r}
library(janitor)
EXP1b_WORK <- clean_names(HO_EXP1b_clean)

```

## Assigning labels to COND levels

```{r}
EXP1b_WORK$cond <- factor(
  EXP1b_WORK$cond,
  levels = c(0, 1),
  labels = c("no origin", "origin")
)
```

## Creating bar plot

```{r}
EXP1b_WORK$cond
as_factor(EXP1b_WORK$cond)

library(ggplot2)
ggplot(EXP1b_WORK, aes(x = as_factor(cond), 
                       fill = as_factor(gender))) +
  geom_bar(position = "dodge")
  
```

## Checking names

```{r}
names(EXP1b_WORK)
```

## Calculate scales

What we need to do:
- reverse coded items?
- Crombach's alpha

```{r}

EXP1b_WORK |>
  select(starts_with("hap")) |>
  names()


class(EXP1b_WORK$cond)
typeof(EXP1b_WORK$cond)
str(EXP1b_WORK$cond)
summary(EXP1b_WORK$cond)

```

## Mapping 

```{r}
item_mapping <- list(
  happiness = c("hap1", "hap2", "hap3"),
  desirablity = c("de1", "de2", "de3")
)
```

## Computing mean for each ppt - each scale separately

```{r}
EXP1b_WORK$hap_mean <- rowMeans(EXP1b_WORK[item_mapping$happiness])
```

## Computing mean for each ppt - all scales at once

```{r}
scale_names <- paste0(names(item_mapping), "_mean")

EXP1b_WORK[scale_names] <- lapply(
  item_mapping,
  function(items) rowMeans(EXP1b_WORK[, items], na.rm = TRUE)
)

# [, column] nothing before ,
# [row,] nothing after ,
```

## Checking the list of variables

```{r}
names(EXP1b_WORK)
```



